An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-07 main.tex
termination proof

We present the rules of \Calculus{} for one automaton in
\cref{tbl:rules:single}. Note that we operate on sets of symbols (terms,
clauses). Additionally, we also use the convention of splitting the formulas
into linear inequalities ($\SomeInequalities$) and any other clauses
($\SomeClause$).

@@

We require that every rule only fires when it would add a new clause. For
example, this means that we cannot use~\Split{} to split on the same term twice.
This suggests a proof strategy where you \Propagate{} when you can, \Split{}
when you must, and \Subsume{} when neither is possible anymore.

\begin{table}
\begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
  Name & \text{Rule} & Prerequisites\\
  \midrule

@@

\subsection{\Calculus{} for one automaton is correct}\label{sec:single:correct}

Since correctness is defined by the formulation of the $\Image{}$~predicate, our
correctness proof amounts to proving inductively that no rule would invalidate
that definition.

\subsubsection{\Calculus{} terminates}
\begin{lemma}\label{lma:single-terminates} A solver implementing the calculus
  described in \cref{sec:calculus} starting with some decidable set of clauses
  $\SomeClause{}$ and an instance of our predicate $\SinglePredicateInstance$
  will eventually terminate.
\end{lemma}

\begin{proof}
The first remaining source of divergence is from leftover predicates. The
$\Image$ predicate can always be trivially removed by applying \Expand{}. This
means a leftover $\Image$ predicate can never cause divergence. The only
remaining predicate then is $\Connected$, which can always be removed by
applying $\Subsume$ when no other of the rules is applicable. Hence all
predicates introduced in the calculus can always be removed.

It remains to prove that each rule is only applied finitely many times. This is
 obvious since both \Propagate{} and \Split{} can only be applied to transition
 variables whose $\Filter$ value has no known lower bound. However, \Split{} can
 always be used to split the proof tree and add clauses in precisely the right
 shape, thereby guaranteeing the ability to make forward progress. Therefore,
 the number of transition variables whose bounds are unknown is decreasing
 monotonically for any given proof goal, and hence termination is guaranteed.
\end{proof}

@@

  \begin{table}[t]
    \begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
      Name & \text{Rule} & Prerequisites\\
      \midrule

>>>>

We present the rules of \Calculus{} for one automaton in
\cref{tbl:rules:single}. Note that we operate on sets of
formulas. Additionally, we also use the convention of splitting the
formulas into linear equations and inequalities ($\SomeInequalities$)
and other formulas ($\SomeClause$). \Fudge{need to say that we assume
  that the predicates (Im, etc.)  only occur positively.}

@@

We require that rules can only be applied when they add new formulas
on every created branch (the notion of \emph{regularity} of a proof is
required~\cite{Fitting96a}). For example, this means that \Split{} can
only be applied to proof goals that contain neither
$\Filter(\Transition) = 0$ nor $\Filter(\Transition) > 0$, and can
never be applied to split on the same term twice on the same branch.
This suggests a proof strategy where you \Propagate{} when you can, \Split{}
when you must, and \Subsume{} when neither is possible anymore.

\begin{table}
\begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
  Name & \text{Rule} & Side conditions\\
  \midrule

@@

\subsection{Correctness of \Calculus{}}\label{sec:single:correct}

Our correctness proof of \Calculus{} consists of two main parts: we
first show that the construction of a proof always terminates, and
then that each of the proof rules in Table~\ref{tbl:rules:single} is
an equivalence transformation, i.e., does not change the set of
satisfying assignments of a formula. In combination, those two results
immediately imply that \Calculus{} gives rise to a decision procedure.

\subsubsection{\Calculus{} terminates}
\begin{lemma}\label{lma:single-terminates}
  Suppose $\SomeClause{}$ is a set of formulas in which the predicates
  $\Image$ and $\Connected$ only occur positively. There is no
  infinite sequence of proofs~$P_0, P_1, P_2, \ldots$ in which $P_0$
  has $\SomeClause{}$ as root, and each $P_{i+1}$ is derived from
  $P_i$ by applying one of the rules in Table~\ref{tbl:rules:single}.
\end{lemma}

\begin{proof}
  The rule~\Expand{} can only be applied finitely often, since each
  application removes one $\Image$ predicate, and none of the rules
  introduce new instances of the predicate. The rule~\Subsume{} can
  only be applied finitely often, since it strictly decreases the
  combined number of $\Image$ and $\Connected$ predicates in sets of
  formulas, and none of the rules increases that number.

  To show termination of \Split{} and \Propagate{}, observe that the
  $\Filter$ in a predicate~$\Connected(\Automaton, \Filter)$ is never
  updated on a proof branch, which means that the set of terms
  $\Filter(t)$ for $t \in \Automaton$ on every branch is finite. Each
  application of \Split{} and \Propagate{} adds a new
  formula~$\Filter(\Transition) = 0$ or $\Filter(\Transition) > 0$
  to a proof goal, which can only happen finitely often.
\end{proof}

@@

  \begin{table}[t]
    \begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
      Name & \text{Rule} & Side conditions\\
      \midrule
